Structuring College Access: The Market Segment Model and College Board Geomarkets

### 2025-05-13 text/back_end_KS.qmd
dallas revisions

The enduring segregated structure of the Chicago metropolitan area from 1980 to 2020 analyzed above plays a significant role in how Geomarkets shape the racial/ethnic and socioeconomic composition of prospect pools. Reflecting the larger metropolitan area dynamics, central Geomarkets that are predominantly comprised of low-income communities of color such as the City of Chicago and South and Southwest Suburbs consistently contribute disproportionately larger shares of Black, Hispanic, and first-generation college prospects, especially at lower SAT score thresholds. On the other hand, predominantly White and affluent Geomarkets in the suburban ring of the metropolitan area like the Chain of Lakes, North Shore, and Evanston and Skokie contribute larger proportions of White and Asian prospects, as well as non-first generation college students. However, the City of Chicago stands out in playing crucial role in fostering college access for first-generation Asian and Hispanic college students, contributing such prospects to the overall pool at significantly higher, disproportionate rates than any other Geomarket. Given the broader racial/ethnic distribution of the Chicago metropolitan area, student list orders that use Geomarket filters excluding the City of Chicago and including more affluent Geomarkets in targeting high-achieving students may still capture some Asian and Hispanic prospects. However, these students are far more likely to be non–first-generation students of color.

@@

@fig-dallas-rq1-race presents the racial and ethnic composition of each Geomarket within the Dallas-Fort Worth metropolitan area across the three census periods. In 1980, all Geomarkets were predominantly White, with White populations ranging from 62% in City of Dallas to nearly 92% in West of Dallas/Fort Worth Metroplex. By 2000, however, Geomarkets located in the central part of the metropolitan area exhibited a more balanced racial/ethnic distribution, including Dallas County (58% White, 16% Black, 19% Hispanic; 6% Asian Pacific Islander), Irving, Arlington, and Grand Prairie (54% White, 12% Black, 26% Hispanic; 16% Asian Pacific Islander), and City of Fort Worth (57% White, 16% Black, 22% Hispanic; 3% Asian Pacific Islander). In contrast, Collin & Rockwall Counties and West of Dallas/Fort Worth Metroplex Geomarkets remained predominantly White in 2000 at 78% and 80%, respectively. Notably, by 2000, the City of Dallas was no longer a predominantly White Geomarket, with a racial/ethnic composition of approximately 24% Black, 30% Hispanic, and 3% Asian Pacific Islander residents.

These shifts in racial and ethnic composition within and across Geomarkets became more pronounced between 2000 and 2020. The Collin and Rockwall and West of Dallas/Fort Worth Metroplex Geomarkets experienced continued declines in their share of White residents. However, both Geomarkets remained predominantly White (58% and 63%, respectively). In contrast, all other Geomarkets experienced continued increases in the shares of Black and Hispanic residents, resulting in a diminished previously White-majority populations for Dallas County (33% White, 23% Black, 32% Hispanic; 9% Asian Pacific Islander), Irving, Arlington, and Grand Prairie (30% White, 20% Black, 36% Hispanic; 12% Asian Pacific Islander), and City of Fort Worth (46% White, 15% Black, 33% Hispanic; 4% Asian Pacific Islander). Lastly, continuous increases in the shares of Black and Hispanic residents further diversified the City of Dallas, making it the most racially/ethnically diverse Geomarket in the Dallas-Fort Worth metropolitan area.

@@

The Dallas-Fort Worth metroplex experienced substantial demographic growth since Geomarket boundaries were dilineated by @RN4982 in the early 1980s. While all six Geomarkets experienced population increases, the most significant growth occurred in the outer suburban areas, reflecting ongoing urban sprawl. Such growth is largely attributed to increases in Hispanic, Black, and Asian populations leading to only two of the six Dallas-Fort Worth Geomarkets remaining predominantly White (Collin and Rockwall Counties and West of Dallas/Fort Worth Metroplex. On the other hand, the metropolitan area has experienced widening disparities in income, poverty, and education attainment across Geomarkets, especially between the historically affluent Collin and Rockwall Counties and the less economically advantaged inner-core Geomarkets (City of Fort Worth and City of Dallas). Together, these trends underscore the growingly racial/ethnic diverse but widening economic inequality within the evolving Geomarket landscape of the Dallas-Fort Worth metroplex.

@@

@fig-dallas-rq2-race presents the racial/ethnic composition of purchased student profiles within Geomarkets in the Dallas-Forthworth area. At the lowest PSAT score thresholds, notable differences emerge in the racial/ethnic composition of prospects within Geomarkets when compared to the overall demographic makeup of the metropolitan area in 2020 (see @fig-dallas-rq1-race). For example, the largest number of prospects in the 1070-1180 PSAT range were from the Collin & Rockwall Counties (57% White, 7% Black, 17% Hispanic, 14% Asian) and West of Dallas/Forth Worth Metroplex Geomarkets (61% White, 7% Black, 19% Hispanic, 7% Asian), both of which yield predominantly White prospect pools. On the other hand, the Dallas County (36% White, 17% Asian, 15% Black, 28% Hispanic) and the Irving, Arlington, and Grand Prairie (34% White, 13% Asian, 15% Black, 34% Hispanic) Geomarket pools have the greatest proportions of non-White prospects (although yielding fewer overall students). Interestingly, while the City of Dallas is the most racially/ethnically diverse Geomarket in the larger metropolitan area, its prospect pool in @fig-dallas-rq2-race includes larger proportion of Hispanic (41%) prospects but lower proportions of Asian (4%) and Black (9%) students comparatively.

These racial/ethnic composition patterns shift as PSAT scores increase. Shares of White prospects within Geomarkets remain relatively proportional to the overall 2020 metropolitan distribution in @fig-dallas-rq1-race. Howeover, prospect pools within Geomarkets from the middle range PSAT score (1190-1260) to the high PSAT score range (1270-1520) become disproportionately more Asian, while the representation of Hispanic and Black prospects decline substantially.

>>>>

The enduring segregated structure of the Chicago metropolitan area from 1980 to 2020 analyzed above plays a significant role in how Geomarkets shape the racial/ethnic and socioeconomic composition of prospect pools. Reflecting the larger metropolitan area dynamics, central Geomarkets that are predominantly comprised of low-income communities of color such as the City of Chicago and South and Southwest Suburbs consistently contribute disproportionately larger shares of Black, Hispanic, and first-generation college prospects, especially at lower SAT score thresholds. On the other hand, predominantly White and affluent Geomarkets in the suburban ring of the metropolitan area like the Chain of Lakes, North Shore, and Evanston and Skokie contribute larger proportions of White and Asian prospects, as well as non-first generation college students. However, the City of Chicago stands out in playing crucial role in fostering college access for first-generation Asian and Hispanic college students, contributing such prospects to the overall pool at significantly higher, disproportionate rates than any other Geomarket. Given the broader racial/ethnic distribution of the Chicago metropolitan area, student list orders that use Geomarket filters excluding the City of Chicago and including more affluent Geomarkets in targeting high-achieving students may still capture some Asian and Hispanic prospects. However, these students are far more unlikely to be first-generation students of color.

@@

@fig-dallas-rq1-race presents the racial and ethnic composition of each Geomarket within the Dallas-Fort Worth metropolitan area across the three census periods. In 1980, all Geomarkets were predominantly White, with White populations ranging from 62% in City of Dallas to nearly 92% in West of Dallas/Fort Worth Metroplex. By 2000, however, Geomarkets located in the central part of the metropolitan area exhibited a more balanced racial/ethnic distribution, including Dallas County (58% White, 16% Black, 19% Hispanic; 6% Asian), Irving, Arlington, and Grand Prairie (54% White, 12% Black, 26% Hispanic; 16% Asian), and City of Fort Worth (57% White, 16% Black, 22% Hispanic; 3% Asian). In contrast, Collin & Rockwall Counties and West of Dallas/Fort Worth Metroplex Geomarkets remained predominantly White in 2000 at 78% and 80%, respectively. Notably, by 2000, the City of Dallas was no longer a predominantly White Geomarket, with a racial/ethnic composition of approximately 24% Black, 30% Hispanic, and 3% Asian residents.

These shifts in racial and ethnic composition within and across Geomarkets became more pronounced between 2000 and 2020. The Collin and Rockwall and West of Dallas/Fort Worth Metroplex Geomarkets experienced continued declines in their share of White residents. However, both Geomarkets remained predominantly White (58% and 63%, respectively). In contrast, all other Geomarkets experienced continued increases in the shares of Black and Hispanic residents, resulting in a diminished previously White-majority populations for Dallas County (33% White, 23% Black, 32% Hispanic; 9% Asian), Irving, Arlington, and Grand Prairie (30% White, 20% Black, 36% Hispanic; 12% Asian), and City of Fort Worth (46% White, 15% Black, 33% Hispanic; 4% Asian). Lastly, continuous increases in the shares of Black and Hispanic residents further diversified the City of Dallas, making it the most racially/ethnically diverse Geomarket in the Dallas-Fort Worth metropolitan area.

@@

The Dallas-Fort Worth metroplex experienced substantial demographic growth since Geomarket boundaries were dilineated by @RN4982 in the early 1980s, leading to an urban landscape with considerable racial/ethnic diversity across multiple Geomarkets (as opposed to concentrated within a single Geomarket like Chicago) but stark socioeconomic disparities between Geomarkets. While all six Geomarkets experienced population increases, the most significant growth occurred in the outer suburban areas, reflecting ongoing urban sprawl. Such growth is largely attributed to increases in Hispanic, Black, and Asian populations leading to only two of the six Dallas-Fort Worth Geomarkets remaining predominantly White (Collin and Rockwall Counties and West of Dallas/Fort Worth Metroplex. On the other hand, the metropolitan area has experienced widening disparities in income, poverty, and education attainment between Geomarkets, especially between the historically affluent Collin and Rockwall Counties and the less economically advantaged inner-core Geomarkets (City of Fort Worth and City of Dallas). Together, these trends underscore the growingly racial/ethnic diverse but widening economic inequality within the evolving Geomarket landscape of the Dallas-Fort Worth metroplex.

@@

@fig-dallas-rq2-race presents the racial/ethnic composition of purchased student profiles within Geomarkets in the Dallas-Forthworth area. At the lowest PSAT score thresholds (1070-1180), notable tiers emerge based on the racial/ethnic composition of prospects between Geomarkets. For instance, the central suburban Geomarkets of Dallas County (36% White, 17% Asian, 15% Black, 28% Hispanic) and Irving, Arlington, and Grand Prairie (34% White, 13% Asian, 15% Black, 34% Hispanic) have the greatest proportions of non-White prospects, which is largely attributed to their greater shares of Asian and Black prospects. While the urban central Geomarkets of City of Dallas and City of Fortworth are the most racially/ethnically diverse in the larger metropolitan area (see @fig-dallas-rq1-race), their prospect pools in @fig-dallas-rq2-race includes larger proportion of Hispanic prospects but lower proportions of Asian and Black students comparatively. The Collin & Rockwall Counties (57% White, 7% Black, 17% Hispanic, 14% Asian) and West of Dallas/Forth Worth Metroplex Geomarkets (61% White, 7% Black, 19% Hispanic, 7% Asian) both yield the largest and predominantly White prospect pools.

<!--# which most closely mirror the overall demographic makeup of the metropolitan area in 2020 (see @fig-dallas-rq1-race). For example,  the Dallas County (36% White, 17% Asian, 15% Black, 28% Hispanic) and the Irving, Arlington, and Grand Prairie (34% White, 13% Asian, 15% Black, 34% Hispanic) Geomarkets have the greatest proportions of non-White prospects although yielding smaller overall pools. Interestingly, while the City of Dallas is the most racially/ethnically diverse Geomarket in the larger metropolitan area, its prospect pool in @fig-dallas-rq2-race includes larger proportion of Hispanic (41%) prospects but lower proportions of Asian (4%) and Black (9%) students comparatively. -->

Prospect pools within Geomarkets from the middle range PSAT score (1190-1260) to the high PSAT score range (1270-1520) become disproportionately more Asian, while the representation of Hispanic and Black prospects decline substantially.
